---
title: "Project1"
format: html
editor: visual
author: Ralph Church
---

### Introduction

The goal of this project is to build reusable functions that interact with the Census Public Use Microdata Sample (PUMS) API. Rather than manually constructing API requests each time, I developed functions that validate user inputs, retrieve data from the API, convert the returned JSON into a tidy tibble, and apply appropriate data types. I also created custom summary and plotting methods for census data objects so that common analyses can be performed consistently across multiple datasets.

```{r}
 # --------------------------
  # imports
  # --------------------------

library(tidyverse)
library(httr)
library(jsonlite)
```

```{r}
# Global variables

census_key <- "c3bbdb3ef53632db1e5b607b6e5b021cff9111af"
```

### Helper Functions

The Census API returns data in JSON format. To make the data easier to work with in R, I first created helper functions that convert API responses into tibbles and recode categorical variables into more meaningful labels. Also a method to read locally kept json files for the time variable look ups.

```{r}
 # --------------------------
  # Helper methods
  # --------------------------

load_time_lookup <- function(file_name) {

  data <- unlist(
    jsonlite::fromJSON(file_name)$values$item
  )

  data
}

recode_factors <- function(data) {

  if ("SEX" %in% names(data)) {
    data$SEX <- factor(
      data$SEX,
      levels = c("1", "2"),
      labels = c("Male", "Female")
    )
  }

  if ("SCH" %in% names(data)) {
    data$SCH <- factor(
      data$SCH,
      levels = c("0", "1", "2", "3"),
      labels = c(
        "Under 3",
        "Not Enrolled",
        "Public School",
        "Private/Home School"
      )
    )
  }

  if ("FER" %in% names(data)) {
    data$FER <- factor(
      data$FER,
      levels = c("0", "1", "2"),
      labels = c(
        "Not Applicable",
        "Yes",
        "No"
      )
    )
  }

  if ("HHL" %in% names(data)) {
    data$HHL <- factor(
      data$HHL,
      levels = c("0", "1", "2", "3", "4", "5"),
      labels = c(
        "Not Applicable",
        "English Only",
        "Spanish",
        "Other Indo-European",
        "Asian/Pacific Island",
        "Other Language"
      )
    )
  }

  if ("SCHL" %in% names(data)) {
    data$SCHL <- factor(
      data$SCHL,
      levels = c(
        "0","01","02","03","04","05","06","07","08","09",
        "10","11","12","13","14","15","16","17","18","19",
        "20","21","22","23","24"
      ),
      labels = c(
        "Under 3",
        "No Schooling",
        "Preschool",
        "Kindergarten",
        "Grade 1",
        "Grade 2",
        "Grade 3",
        "Grade 4",
        "Grade 5",
        "Grade 6",
        "Grade 7",
        "Grade 8",
        "Grade 9",
        "Grade 10",
        "Grade 11",
        "12th No Diploma",
        "High School Diploma",
        "GED",
        "Some College <1 Year",
        "College Credit No Degree",
        "Associate Degree",
        "Bachelor Degree",
        "Master Degree",
        "Professional Degree",
        "Doctorate Degree"
      )
    )
  }

  data
}

# Census encodes JWAP/JWDP as time intervals.
# These are converted to time values and are not
# included in summary.census().
convert_time_variable <- function(x, lookup) {

  x <- as.numeric(x)

  x[x == 0] <- NA

  result <- lookup[x]

  result <- sapply(
    result,
    function(interval) {

      if (is.na(interval)) {
        return(NA_character_)
      }

      start_time <- strsplit(interval, " to ")[[1]][1]

      start_time <- gsub("\\.", "", start_time)

      start <- strptime(
        start_time,
        format = "%I:%M %p"
      )

      midpoint <- start + 120

      format(
        midpoint,
        "%I:%M %p"
      )
    }
  )

  unname(result)
}

```
First, we load the time variable json files in order to more easily convert:

```{r}
jwap_lookup <- load_time_lookup("JWAP.json")
class(jwap_lookup)

str(jwap_lookup)
jwdp_lookup <- load_time_lookup("JWDP.json")
```

The parse_pums() function extracts the JSON response, converts it into an R object, and transforms the returned data into a tidy tibble with appropriate column names.

```{r}
parse_pums <- function(response) {

  # Extract the API response as text
  response_text <- content(response, "text")

  # Convert the JSON response into an R object
  parsed <- fromJSON(response_text)

  # Remove the first row and convert the remaining data to a tibble
  pums_data <- as_tibble(parsed[-1, ])

  # Use the first row as the column names
  colnames(pums_data) <- parsed[1, ]

  # Return the cleaned tibble
  pums_data
}
```

### Querying the Census API

The primary function developed for this project is query_pums(). This function allows a user to specify a survey year, numeric variables, categorical variables, geography level, and geographic subset. Input validation is performed to ensure that only supported years and variables are requested. The function also guarantees that the survey weight variable (PWGTP) is always included.

```{r}
query_pums <- function(
    year = 2022,
    numeric_vars = c("AGEP", "PWGTP"),
    categorical_vars = c("SEX"),
    geography = "state",
    subset_value = "37", # North Carolina by default
    key
) {

  # --------------------------
  # Validate year
  # --------------------------

  if (!is.numeric(year) || length(year) != 1) {
    stop("year must be a single numeric value.")
  }

  if (!(year %in% 2010:2022)) {
    stop("year must be between 2010 and 2022.")
  }

  # --------------------------
  # Validate numeric variables
  # --------------------------

  valid_numeric <- c(
    "AGEP",
    "GASP",
    "GRPIP",
    "JWAP",
    "JWDP",
    "JWMNP",
    "PWGTP"
  )

  if (!all(numeric_vars %in% valid_numeric)) {
    stop("One or more numeric variables are invalid.")
  }

  # Ensure PWGTP is always returned
  if (!("PWGTP" %in% numeric_vars)) {
    numeric_vars <- c(numeric_vars, "PWGTP")
  }

  # Require at least one numeric variable besides PWGTP
  if (length(setdiff(numeric_vars, "PWGTP")) < 1) {
    stop("At least one numeric variable besides PWGTP must be selected.")
  }

  # --------------------------
  # Validate categorical variables
  # --------------------------

  valid_categorical <- c(
    "FER",
    "HHL",
    "SCH",
    "SCHL",
    "SEX"
  )

  if (!all(categorical_vars %in% valid_categorical)) {
    stop("One or more categorical variables are invalid.")
  }

  if (length(categorical_vars) < 1) {
    stop("At least one categorical variable must be selected.")
  }

  # --------------------------
  # Validate geography
  # --------------------------

  geography <- tolower(geography)

  valid_geography <- c(
    "region",
    "division",
    "state"
  )

  if (!(geography %in% valid_geography)) {
    stop("geography must be region, division, or state.")
  }
  
  if (geography == "region" &&
    subset_value != "all" &&
    !(subset_value %in% c("1", "2", "3", "4"))) {
    stop("Region must be 1, 2, 3, or 4.")
  }
  
  if (geography == "division" &&
      subset_value != "all" &&
      !(subset_value %in% as.character(1:9))) {
    stop("Division must be between 1 and 9.")
  }

  # --------------------------
  # Build variable list
  # --------------------------

  variables <- unique(
    c(
      numeric_vars,
      categorical_vars
    )
  )

  variable_string <- paste(
    variables,
    collapse = ","
  )

  # --------------------------
  # Geography filter
  # --------------------------

  geography_clause <- paste0(
    "&for=",
    geography,
    ":",
    subset_value
  )

  # --------------------------
  # Build URL
  # --------------------------

  url <- paste0(
    "https://api.census.gov/data/",
    year,
    "/acs/acs1/pums?",
    "get=",
    variable_string,
    geography_clause,
    "&key=",
    key
  )

  # --------------------------
  # Query API
  # --------------------------

  response <- GET(url)

  # --------------------------
  # Convert to tibble
  # --------------------------
  
  # Convert response to tibble
  data <- parse_pums(response)
  
  numeric_columns <- intersect(
  names(data),
  c(
    "AGEP",
    "GASP",
    "GRPIP",
    "JWMNP",
    "PWGTP"
  )
)

data <- data |>
  mutate(
    across(
      all_of(numeric_columns),
      as.numeric
    )
  )

# Convert JWAP if requested
if ("JWAP" %in% names(data)) {
  data$JWAP <- convert_time_variable(
    data$JWAP,
    jwap_lookup
  )
}

# Convert JWDP if requested
if ("JWDP" %in% names(data)) {
  data$JWDP <- convert_time_variable(
    data$JWDP,
    jwdp_lookup
  )
}

factor_columns <- intersect(
  names(data),
  c(
    "FER",
    "HHL",
    "SCH",
    "SCHL",
    "SEX"
  )
)

data <- data |>
  mutate(
    across(
      all_of(factor_columns),
      as.factor
    )
  )

  # make the SEX variable more readable by assigning Male and Female instead of 1 and 2
  data <- recode_factors(data)

  # Add custom census class
  class(data) <- c("census", class(data))

  return(data)
}
```

The following examples demonstrate that the function successfully retrieves data using default parameters, custom variable selections, and automatic inclusion of the required weight variable.

### Unit Tests

Test 1: the defaults:

```{r}
query_pums(
  key = census_key
)
```

Test 2: Different year and variables:

```{r}
query_pums(
  year = 2018,
  numeric_vars = c("AGEP", "GRPIP"),
  categorical_vars = c("SEX", "SCHL"),
  key = census_key
)
```

Test 3: PWGTP automatically included:

```{r}
query_pums(
  numeric_vars = c("AGEP"),
  key = census_key
)
```

Test 4: Querying with JWAP:

```{r}
test_jwap <- query_pums(
  numeric_vars = c("JWAP"),
  key = census_key
) |>
  pull(JWAP)

str(test_jwap)

head(test_jwap, 20)
```

The Census API stores **JWAP** and **JWDP** as coded time intervals rather than actual clock times. The helper function uses the lookup table to convert each code into its corresponding time interval and then returns the midpoint of that interval. For example, the interval "3:15 p.m. to 3:19 p.m." is represented as "03:17 PM". Missing values (coded as 0) are converted to NA.

Test 5 Querying with JWDP: 

```{r}
query_pums(
  numeric_vars = c("JWDP"),
  key = census_key
) |>
  pull(JWDP) |>
  head(10)
```


Test 6: FER recoding:

```{r}
query_pums(
  categorical_vars = c("FER"),
  key = census_key
) |>
  count(FER)
```

Test 7: HHL Recoding:

```{r}
query_pums(
  categorical_vars = c("HHL"),
  key = census_key
) |>
  count(HHL)
```

Test 8: SCHL Recoding:

```{r}
query_pums(
  categorical_vars = c("SCHL"),
  key = census_key
) |>
  count(SCHL)
```

Test 9: Querying the region:

```{r}
query_pums(
  geography = "region",
  subset_value = "3", # Region 3 = South
  key = census_key
)
```

Test 10: Querying Division:

```{r}
query_pums(
  geography = "division",
  subset_value = "5", # Division 5 = South Atlantic
  key = census_key
)
```

### Combining Multiple Years

Because many analyses require multiple years of data, I created a wrapper function that repeatedly calls query_pums() and combines the results into a single tibble. The function excludes the year 2020 because the Census Bureau advises against using those data due to disruptions caused by the COVID-19 pandemic.

```{r}
query_pums_years <- function(
    years,
    numeric_vars = c("AGEP", "PWGTP"),
    categorical_vars = c("SEX"),
    geography = "state",
    subset_value = "37", # North Carolina by default
    key
) {

  # Remove 2020
  if (2020 %in% years) {
    stop("2020 is not allowed because Census discarded that data.")
  }

  results <- lapply(
    years,
    function(y) {

      query_pums(
        year = y,
        numeric_vars = numeric_vars,
        categorical_vars = categorical_vars,
        geography = geography,
        subset_value = subset_value,
        key = key
      ) |>
        mutate(year = y)

    }
  )

  bind_rows(results)

}
```

Now, we make a call in order to test the multi-year function:

```{r}
query_pums_years(
  years = c(2018, 2019, 2021, 2022),
  key = census_key
)
```

We now check the class of our object to ensure it has the correct properties:

```{r}
my_data <- query_pums(key = census_key)

class(my_data)
```

### Custom Census Class

To support custom methods, the returned tibble is assigned an additional class called "census". This allows generic functions such as summary() and plot() to automatically use methods specifically designed for census data.

A custom summary method was created for census objects. Numeric variables are summarized using weighted means and weighted standard deviations based on the Census survey weight (PWGTP). Categorical variables are summarized using frequency counts. By default, all eligible variables are summarized, although the user may specify a subset of variables if desired.

```{r}
summary.census <- function(
    object,
    numeric_vars = NULL,
    categorical_vars = NULL,
    ...
) {

  if (is.null(numeric_vars)) {

    numeric_vars <- names(object)[sapply(object, is.numeric)]

    # JWAP and JWDP are Census-coded time intervals.
    # These variables are excluded from summary statistics
    # because the project instructions allow time variables
    # to be skipped.
    numeric_vars <- setdiff(
      numeric_vars,
      c(
        "PWGTP",
        "year",
        "JWAP",
        "JWDP"
      )
    )

  }

  if (is.null(categorical_vars)) {

    categorical_vars <- names(object)[sapply(object, is.factor)]

  }

  results <- list()

  # Weighted summaries
  for (var in numeric_vars) {

    x <- object[[var]]
    w <- object$PWGTP

    # Handle missing values if any occur.
    weighted_mean <- sum(x * w, na.rm = TRUE) /
      sum(w[!is.na(x)])
    
    weighted_sd <- sqrt(
      sum(x^2 * w, na.rm = TRUE) /
        sum(w[!is.na(x)]) -
        weighted_mean^2
    )

    results[[var]] <- list(
      mean = weighted_mean,
      sd = weighted_sd
    )

  }

  # Categorical counts
  for (var in categorical_vars) {

    results[[var]] <- table(
      object[[var]]
    )

  }

  return(results)

}

my_data <- query_pums(key = census_key)

str(my_data)
summary(my_data) # will now automatically call summary.census() since first class is "census" printed out previously using class()
```

The results show weighted summary statistics for numeric variables and frequency distributions for categorical variables, providing a quick overview of the returned data.

### Census Plot Method

A custom plotting method was also developed for census objects. The user specifies a categorical variable and a numeric variable, and the function produces a weighted boxplot using the Census survey weights. This provides a simple way to compare distributions across demographic groups.

```{r}
plot.census <- function(
    x,
    cat_var,
    num_var,
    ...
) {

  ggplot(
    x,
    aes(
      x = get(cat_var),
      y = get(num_var),
      weight = PWGTP
    )
  ) +
    geom_boxplot() +
    labs(
      x = cat_var,
      y = num_var,
      title = paste(
        num_var,
        "by",
        cat_var
      )
    )

}
```

The boxplot below compares age distributions between males and females in North Carolina using survey-weighted observations.

```{r}
plot(
  my_data,
  cat_var = "SEX",
  num_var = "AGEP"
)
```

The second example demonstrates that the plotting function can be reused with different variables without modification.

```{r}
plot(
  query_pums(
    numeric_vars = c("GRPIP"),
    categorical_vars = c("SEX"),
    key = census_key
  ),
  cat_var = "SEX",
  num_var = "GRPIP"
)
```

GRPIP (Gross Rent as a Percentage of Household Income) is heavily right-skewed as is shown in the figure above.

### Exploring the Data

Example: Age and School Enrollment

To demonstrate the functions developed in this project, I explored the relationship between age (AGEP) and school enrollment status (SCH). The results show that people currently attending school are generally younger than those who are not enrolled. This example illustrates how the API query, summary, and plotting functions can be combined to quickly investigate patterns in Census data.

```{r}
edu_data <- query_pums(
  numeric_vars = c("AGEP"),
  categorical_vars = c("SCH"),
  key = census_key
)

summary(
  edu_data,
  numeric_vars = "AGEP",
  categorical_vars = "SCH"
)

plot(
  edu_data,
  cat_var = "SCH",
  num_var = "AGEP"
)
```

People who are currently enrolled in school tend to be younger, while people who are not enrolled tend to be older. This is exactly what we would expect and provides a good example of how the custom summary and plotting functions can be used to explore Census data.